---
title: "Demo functions"
author: "Marc A.T. Teunis"
date: "4/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages
```{r}
library(tidyverse)
```

## Data
```{r}
data <- tibble(
  a = rnorm(100, 1, 0.5),
  b = rnorm(100, 2, 0.2),
  c = rnorm(100, 3, 0.7)
)

data

```

## Opdracht
Normaliseer iedere kolom:

$\frac{Yi - min(Y)}{max(Y) - min(Y)}$

```{r, eval=FALSE}
data$a_rescaled <- (data$a - min(data$a)) / (max(data$a) - min(data$a))
data
data$brescaled <- (data$b - min(data$b)) / (max(data$b) - min(data$b))
data$c_rescaled <- (data$c - min(data$c)) / (max(data$c) - min(data$c))

data


```

```{r}
## vector meestal aangeven met een x
## dummie 
x = c(1:10)

rescale <- function(x){
  
    min_value <- min(x, na.rm = TRUE)
    max_value <- max(x, na.rm = TRUE)
    range_value <- max_value - min_value
    normalized_x <- (x - min_value) / (max_value - min_value)
  return(normalized_x)
  
}


# data$a_rescaled <- rescale(x = data$a)
# data$b_rescaled <- rescale(x = data$b)
# data$c_rescaled <- rescale(x = data$c)

data_rescaled <- map_df(
  .x = data,
  .f = rescale
)

data_rescaled
```






## Normalize (scale)
Normalize each column for min max according $\frac{Yi - Y.min}{Ymax - Ymin}$
value between 0 and 1
See for a [free online LaTeX editor:](https://www.codecogs.com/latex/eqneditor.php

## What happens if vector contains an NA?


## Write a function that nromalizes a single column

## use dot dot dot (ellipsis)

## Use x = NULL to force user to address argument

## Show iteration with `map()` from 
