---
title: "Demo functions"
author: "Marc A.T. Teunis"
date: "4/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages
```{r}
library(tidyverse)
```

## Data
```{r}
data <- tibble(
  a = rnorm(100, 1, 0.5),
  b = rnorm(100, 2, 0.2),
  c = rnorm(100, 3, 0.7)
)

data

```

## Normalize (scale)
Normalize each column for min max according $\frac{Yi - Y.min}{Ymax - Ymin}$
value between 0 and 1
See for a [free online LaTeX editor:](https://www.codecogs.com/latex/eqneditor.php

## Opdracht
Normaliseer iedere kolom in dataframe `data`:

$\frac{Yi - min(Y)}{max(Y) - min(Y)}$

```{r, eval=FALSE}
data$a_rescaled <- (data$a - min(data$a)) / (max(data$a) - min(data$a))
data
data$brescaled <- (data$b - min(data$b)) / (max(data$b) - min(data$b))
data$c_rescaled <- (data$c - min(data$c)) / (max(data$c) - min(data$c))

data


```

## A function that reduces duplication in the code
```{r}
## vector in R as a function argument is usually indicated as x

## dummy argument to test the body of the function
x = c(1:10)

rescale <- function(x){
  
    min_value <- min(x)
    max_value <- max(x)
    range_value <- max_value - min_value
    normalized_x <- (x - min_value) / (max_value - min_value)
  return(normalized_x)
  
}

## function tests:
# data$a_rescaled <- rescale(x = data$a)
# data$b_rescaled <- rescale(x = data$b)
# data$c_rescaled <- rescale(x = data$c)

```

## Iterate function over `data` columns
```{r}
data_rescaled <- map_df(
  .x = data,
  .f = rescale
)

data_rescaled
```

## What happens if vector contains an NA?
```{r}
## test function with vector containing an NA
with_na <- c(1:9, NA)
rescale(with_na)

## vector meestal aangeven met een x
## dummie 
x = c(1:10)

## fix for missing values
rescale <- function(x){
  
  min_value <- min(x)
  max_value <- max(x, na.rm = TRUE)
  range_value <- max_value - min_value
  normalized_x <- (x - min_value) / (max_value - min_value)
  
  return(normalized_x)
  
}


```

## Use `dot dot dot` (ellipsis) argument
```{r}
## test function with vector containing an NA
with_na <- c(1:9, NA)
rescale(with_na)

## vector meestal aangeven met een x
## dummie 
x = c(1:10, NA)

## fix for missing values
rescale <- function(x, ...){
  
  min_value <- min(x, ...)
  max_value <- max(x, ...)
  range_value <- max_value - min_value
  normalized_x <- (x - min_value) / (max_value - min_value)
  
  return(normalized_x)
  
}

rescale(x = x, na.rm = TRUE)
```


## Use x = NULL to force user to address argument
```{r}
## test function with vector containing an NA
with_na <- c(1:9, NA)
rescale(with_na)

## vector meestal aangeven met een x
## dummie 
x = c(1:10)

## fix for missing values
rescale <- function(x = NULL, ...){
  
  min_value <- min(x, ...)
  max_value <- max(x, ...)
  range_value <- max_value - min_value
  normalized_x <- (x - min_value) / (max_value - min_value)
  
  return(normalized_x)
  
}

rescale(x = x, na.rm = TRUE)
```

## Function documentation using `roxygen2`
In order to standardize documentation for functions in R we can use roxygen coments. These are special comments and operators (`@operator`) to build the documentation. Make it a habit to write this documentation early in the process of developing a function and come back to it if things have changed.
```{r}
#' @title Normalize a numeric vector. 
#' @param x A numeric vector of type double or integer, or coercible to one.
#' @param ... Additional arguments passed on to min() and max().
#' @return Normalized vector according the formula 
#' \eqn{\frac{Yi - min(Y)}{max(Y) - min(Y)}} where Yi is the i-th element in #' the vector x. The dot dot dot (elippsis) argument can be set to handle 
#' missing values. Set na.rm = TRUE in the function call. Default, missing 
#' values are not handled.
#' @details The process of nromalization usually reduces the scale on which the data is represented. In this case a often used function is used that reduces the scale of the data to a respresentation between 0 and 1, in which the maximum value, the minumum vale are used as parameters. Other way of normalization use e.g. the standard deviation, the mean or the variance as parameters. 
#' @example 
#' x <- c(1:10)
#' rescale(x)
#' z <- c(1:9, NA)
#' rescale(z, na.rm = TRUE)
#' @export

rescale <- function(x = NULL, ...){
  
  min_value <- min(x, ...)
  max_value <- max(x, ...)
  range_value <- max_value - min_value
  normalized_x <- (x - min_value) / (max_value - min_value)
  
  return(normalized_x)
  
}


```

